---
title: "NFL Win Probability"
output: 
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: scroll
    theme: cosmo
    social: [ "twitter", "linkedin", "menu" ]
runtime: shiny
---

<style>                     
.navbar {
  background-color:#000000;
  border-color:white;
}
.navbar-brand {
color:white!important;
}
</style>   

<style type="text/css">

.chart-title {  /* chart_title  */
   font-size: 15px

</style>


```{r global, include = FALSE}

library(flexdashboard)
library(shiny)
library(plotly)
library(reactable)
library(reactablefmtr)
library(nflfastR)
library(nflseedR)
library(tidyverse)
library(teamcolors) # NFL team colors and logos
library(extrafont)
library(gt)
library(gtExtras)
library(ggimage)
library(animation)
library(DBI)
library(RSQLite)
library(glue)
library(ggtext)

# Optional but makes R prefer not to display numbers in scientific notation
options(scipen = 9999)

# Custom ggplot theme (inspired by Owen Phillips at the F5 substack blog)
theme_custom <- function () { 
  theme_minimal(base_size=11, base_family="Chivo") %+replace% 
    theme(
      panel.grid.minor = element_blank(),
      plot.background = element_rect(fill = 'transparent', color = "transparent")
    )
}


# Define 538 table theme for Reactable table(s) below
theme_538 <- function() {
    reactable::reactableTheme(
        searchInputStyle = list(width = "31%", backgroundColor = "#F9F9F9"),
        style = list(
            fontFamily = "Chivo"
        ),
        headerStyle = list(
            "&:hover[aria-sort]" = list(
                background = "hsl(0, 0%, 80%)"),
            "&[aria-sort='ascending'], &[aria-sort='descending']" = list(
                background = "#555",
                color = "#FFF"
            ),
            borderColor = "#333"
        ),
        borderColor = "#CDCDCD"
    )
}


# load play by play data from nflfastR
pbp_data <- load_pbp(2022)

#Load games data from Lee Sharpe/nflseedR
games <- load_sharpe_games() %>% 
  filter(season == 2022)


```

Inputs {.sidebar}
-----------------------------------------------------------------------

``` {r}

 selectInput("gameInput", label = h3("Select Game"), 
    choices = unique(pbp_data$game_id),
    selected="2022_01_BUF_LA", multiple = FALSE)

```

Choose a game from the 2022 season to see its play-by-play win probability chart. For example, selecting "2022_01_BUF_LA" will show you win probability for the 2022 Week 1 matchup between the Buffalo Bills and Los Angeles Rams.

Row
-----------------------------------------------------------------------

### Win Probability Charts
    
```{r}

# select game id and filter data
 wp_plot <- reactive({
        pbp_data %>%
     filter(game_id == input$gameInput) %>%
     filter(!is.na(home_wp),
            !is.na(away_wp)) %>%
     select(game_seconds_remaining,
                    home_wp, away_wp) %>%
  gather(team, wpa, -game_seconds_remaining)
    })


 # make wp plot
renderPlotly({
  ggplotly(ggplot(wp_plot(), aes(x = game_seconds_remaining, y = wpa, color = team)) +
  geom_line(size = 1.2) +
  geom_hline(yintercept = 0.5, color = "gray", linetype = "dashed") +
  #scale_color_manual(labels = c("SF", "GB"),
   #                  values = c(la_color, tb_color),
    #                 guide = FALSE) +
  #scale_x_reverse(breaks = seq(0, 3600, 300)) + 
  geom_vline(xintercept = 900, linetype = "dashed") + 
  geom_vline(xintercept = 1800, linetype = "dashed") + 
  geom_vline(xintercept = 2700, linetype = "dashed") + 
  geom_vline(xintercept = 0, linetype = "dashed") + 
  labs(
    x = "",
    y = "",
    title = "2022 Win Probability",
    subtitle = glue("Team Home vs Team Away")
  ) + 
  theme_custom() +
  theme(plot.title = element_text(face="bold")) +
  scale_x_continuous(
    trans = "reverse",
    breaks = c(2700, 1800, 900, 0), 
    labels = c("END\nQ1", "HALF\nTIME", "END\nQ3", "END\nQ4")) +
  scale_y_continuous(labels = scales::percent_format()) +
  theme(plot.subtitle = element_markdown())
  )
})

```


Row {.tabset}
-----------------------------------------------------------------------

### Games
    
```{r}

# Make games table
games2 <- games %>% 
  select(game_id, game_type, gameday, week, away_team:home_score, spread_line, 
         away_spread_odds, home_spread_odds,away_qb_name, home_qb_name, stadium)

# Get team logos
team_logos <- nflfastR::teams_colors_logos

# Get player headshots
rosters <- nflfastR::fast_scraper_roster(seasons = 2022) %>% 
  select(full_name, headshot_url)


# join in team logos and headshots for inclusion in table
games2 <- left_join(games2, team_logos, by = c("away_team" = "team_abbr")) #Away teams
games2 <- left_join(games2, team_logos, by = c("home_team" = "team_abbr")) #Home teams

#games2 <- left_join(games2, rosters, by = c("away_qb_name" = "full_name")) #Away QBs
#games2 <- left_join(games2, rosters, by = c("home_qb_name" = "full_name")) #Home QBs



# Reactable table setup

games_table <- reactive({
  games2 %>%
    select(week, gameday, game_type, team_logo_espn.x, away_team, away_score, team_logo_espn.y, home_team, home_score, spread_line,  away_spread_odds, home_spread_odds) %>% 
  rename(team_logo_espn_away = team_logo_espn.x,
         team_logo_espn_home = team_logo_espn.y)
    })


# render table
renderReactable({
  reactable(games_table(),
  theme = theme_538,
                  showSortIcon = TRUE,
                  searchable = TRUE,
                  language = reactableLang(
                      searchPlaceholder = "SEARCH FOR A TEAM (ABBR)..."),
                  defaultPageSize = 100,
                  columns = list(
                      week = colDef(name = "Week",
                                       align = "center"),
                      gameday = colDef(name = "Date",
                                       align = "right",
                                       minWidth = 120),
                      game_type = colDef(name = "Round",
                                         style = list(borderRight = "2px solid #555"),
                                       align = "right"),
                      away_team = colDef(name = "Away",
                                       align = "right"),
                      away_score = colDef(name = "Away Score",
                                       align = "right"),
                      home_team = colDef(name = "Home",
                                       align = "right"),
                      home_score = colDef(name = "Home Score",
                                       align = "right"),
                      spread_line = colDef(name = "Spread Line",
                                           style = list(borderLeft = "2px solid #555"),
                                       align = "right"),
                      away_spread_odds = colDef(name = "Away Spread",
                                       align = "right"),
                      home_spread_odds = colDef(name = "Home Spread",
                                       align = "right"),
                      
                      ### add logos using embed_img()
                      team_logo_espn_home = colDef(
                          name = "",
                          maxWidth = 40,
                          align = "right",
                          cell = embed_img(height = "25", width = "30")
                      ),
                      team_logo_espn_away = colDef(
                          name = "",
                          maxWidth = 40,
                          align = "right",
                          cell = embed_img(height = "25", width = "30")
                      )),
  
                  pagination = FALSE,
                  compact = TRUE, 
                  borderless = FALSE, 
                  striped = FALSE,
                  fullWidth = FALSE, 
                  defaultColDef = colDef(align = "center", minWidth = 95)
        ) %>% 
  add_title("2022 NFL Season",
            font_size = 28) %>% 
  add_subtitle("Every NFL game by week, including scores and Vegas spreads. Click a column header to sort by that statistic.",
               font_size = 14,
               font_weight = "normal") %>% 
  add_source("Data: nflverse")
  
})
 

```


### About

**NFL Win Probability Tool**

This dashboard looks at individual game win probability. It uses play-by-play data from Ben Baldwin and Sebastian Carl’s nflfastR package.  nflfastR is an extension of the original work done by the nflscrapR team (Maksim Horowitz, Ron Yurko, and Sam Ventura). App created by [Stephan Teodosescu](https://stephanteodosescu.com/).


More on these metrics and the models underpinning them can be found at [Open Source Football](https://www.opensourcefootball.com/posts/2020-09-28-nflfastr-ep-wp-and-cp-models/).


This dashboard is built with R using the [Rmarkdown](https://rmarkdown.rstudio.com/) and [flexdashboard](https://rmarkdown.rstudio.com/flexdashboard/) frameworks. The code behind the dashboard is made available [here](https://github.com/steodose/NFL-win-probability)


**Other Info**

Check out my [NFL Analytics site](https://stephanteodosescu.com/NFL/NFL-Summary-Report.html) for more advanced stats and analysis of the current NFL season. For any questions or feedback, you can contact me on [Twitter](https://twitter.com/steodosescu).
